---
title: "Estimating aphid population parameters"
author: "Lucas A. Nell"
date: "2018-07-03"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Estimating aphid population parameters}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "svglite", 
  fig.width = 6,
  fig.height = 4
)
knitr::opts_knit$set(root.dir = normalizePath(".."))
options(tibble.print_min = 4L, tibble.print_max = 8L)
```
```{r source_Rprofile, echo = FALSE}
if (!isTRUE(getOption('knitr.in.progress'))) source(".Rprofile")
```

```{r library_pkgs, warning=FALSE, echo=FALSE}
suppressPackageStartupMessages({
    library(clonewars)
})
bayesplot::color_scheme_set("viridis")
```


This file will eventually contain the analyses to estimate aphid population-growth
parameters.
It is not yet finished, and much of the code below is not run during the making of
this document.


## Imputing missing data

See `?impute` for information on imputing missing data.


## Load data

```{r load_data}
growth <- 
    load_data() %>%
    # Now filter out early part of each time series, before N > 6
    # N <= 6 is when the stochasticity associated with only starting with 2 adults
    # appears to be strongest
    group_by(line, rep) %>% 
    filter(1:n() >= which(N > 6)[1]) %>%
    mutate(date = date - min(date)) %>%
    ungroup()
```

This is how many reps I have of each line:

```{r view_reps, echo=FALSE}
growth %>% 
    group_by(line) %>% 
    summarize(n = rep %>% unique() %>% length()) %>% 
    knitr::kable()
```




## See time series data

```{r filter_data}
growth %>% 
    group_by(line) %>% 
    mutate(rep = map_int(rep, ~ which(sort(unique(rep)) == .x))) %>% 
    ungroup() %>% 
    mutate(rep = factor(rep)) %>% 
    ggplot(aes(date, X, color = rep)) +
    geom_line(size = 0.75) +
    geom_point(size = 0.75) +
    facet_wrap(~ line, nrow = 2) +
    scale_color_brewer(palette = "RdBu") +
    NULL
```






## Convert data for stan

```{r convert_data}
model_data <- growth %>% 
    line_data()
str(model_data, give.attr = FALSE)
```


## Run model

```{r run_model, eval = FALSE}
# Takes ~ 42 sec at adapt_delta = 0.9
set.seed(65136)
stan_fit <- model_data %>%
    fit_lines(
        control = list(adapt_delta = 0.9)
    )
```
```{r write_model, eval = FALSE, echo = FALSE}
write_rds(stan_fit, "data-raw/stan_fit.rds")
```
```{r read_model, eval = TRUE, echo = FALSE}
stan_fit <- read_rds("data-raw/stan_fit.rds")
```


## Examine model output


```{r examine_model, message=FALSE}

# http://mc-stan.org/bayesplot/articles/visual-mcmc-diagnostics.html

posterior <- as.array(stan_fit)

mcmc_intervals(posterior, pars = sprintf("R[%i]", 1:8)) +
    scale_y_discrete(labels = growth$line %>% levels()) +
    ggtitle("Growth rates")
mcmc_intervals(posterior, pars = sprintf("A[%i]", 1:8)) +
    scale_y_discrete(labels = growth$line %>% levels()) +
    ggtitle("Density dependences")
```


This code is looking at model output. I'm not including it yet in this vignette because
it's not ready.


```{r examine_model2, eval = FALSE}
apply(rstan::extract(stan_fit, "R", permuted = FALSE), 3, mean)
apply(rstan::extract(stan_fit, "A", permuted = FALSE), 3, mean)

apply(rstan::extract(stan_fit, "s_epsilon", permuted = FALSE), 3, mean)
apply(rstan::extract(stan_fit, "s_epsilon", permuted = FALSE), 3, quantile,
      probs = c(0.025, 0.975))



print(stan_fit, digits = 6, pars = c("R", "A", "s_epsilon"), probs = c())

# parameters of interest
pars_interest <- c(sprintf("A[%i]", 1:8), sprintf("R[%i]", 1:8),
                   paste0(rep(c("", "s_"),each=2), c("theta", "phi")),
                   "hats_phi", "s_epsilon")

# Rhats
rhats <- rhat(stan_fit, pars = pars_interest)
mcmc_rhat(rhats)  + yaxis_text(hjust = 1)

# Neff / N
neff_ratios <- neff_ratio(stan_fit, pars = pars_interest)
mcmc_neff(neff_ratios, size = 2) + yaxis_text(hjust = 1)


# Autocorrelation
mcmc_acf(posterior, pars = pars_interest[1:8], lags = 10)
mcmc_acf(posterior, pars = pars_interest[9:16], lags = 10)
mcmc_acf(posterior, pars = pars_interest[17:length(pars_interest)], lags = 10)


lp <- log_posterior(stan_fit)
np <- nuts_params(stan_fit)

mcmc_trace(posterior, pars = pars_interest[1:8], np = np)
mcmc_trace(posterior, pars = pars_interest[9:16], np = np)
mcmc_trace(posterior, pars = pars_interest[17:length(pars_interest)], np = np)

mcmc_nuts_divergence(np, lp)

mcmc_parcoord(posterior, np = np)

mcmc_scatter(
  posterior,
  pars = c("s_epsilon", "R[6]"),
  transform = list(`R[6]` = "log"),
  np = np
)


growth %>% 
    group_by(line) %>% 
    summarize(n = length(unique(rep))) %>% 
    mutate(line = as.integer(line)) %>% 
    as.data.frame()


pred_df <- make_pred_df(stan_fit, growth)

pred_df %>%
    # filter(rep == 1) %>% 
    # filter(rep == 4, line == "Clover-2017-2") %>%
    ggplot(aes(date, X)) +
    geom_ribbon(aes(ymin = X_lower, ymax = X_upper),
                fill = "gray70", alpha = 0.5) +
    geom_line(color = "firebrick", alpha = 0.5) +
    geom_point(shape = 1, color = "firebrick", alpha = 0.5) +
    geom_line(aes(y = X_pred), color = "dodgerblue3", size = 0.5) +
    # facet_wrap(~ rep + line, nrow = 2) +
    facet_wrap(~ rep + line, nrow = 4) +
    NULL



```


