---
title: "Choose priors"
author: "Lucas A. Nell"
date: "2018-07-04"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choose priors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "svglite", 
  fig.width = 6,
  fig.height = 4
)
knitr::opts_knit$set(root.dir = normalizePath(".."))
# source(".Rprofile")
bayesplot::color_scheme_set("viridis")
```


```{r library_pkgs, warning=FALSE, echo=FALSE}
suppressPackageStartupMessages({
    library(aphidreps)
})
```


## Model

Below is the model I'm fitting:

$$
X_{t+1} = X_t + R_0 \left( 1 - \alpha ~ \text{e}^{X_t} \right)
$$

where $X_t$ is log aphid counts for a particular plant at time $t$,
$R_0$ is the aphid line's growth rate, and $\alpha$ is the line's density dependence.
Growth rates get one parameter for each aphid line.
Density dependence, however, gets both mean and standard deviation estimates per aphid
line, plus an overall $\alpha$ estimate per time-series, as a sort of random effect.

For clonal line $i$ on plant $j$, $\alpha$ is calculated as follows:

$$
\alpha_{i,j} = \bar{\alpha} + Z_{\alpha,i} ~ \sigma_{\alpha} + 
    {\hat{Z}}_{\alpha,j} ~ {\hat{\sigma}}_{\alpha,i}
$$

Here,
$\bar\alpha$ is the mean $\alpha$ among all lines,
$Z_{\alpha,i}$ is a randomly generated z-score for clonal line $i$,
$\sigma_\alpha$ is for among-line variability in $\alpha$,
${\hat{Z}}_{\alpha,j}$ is a randomly generated z-score for plant $j$,
${\hat\sigma}_{\alpha,i}$ is for within-line variability in $\alpha$ for line $i$.


> Note that all parameters below are truncated to be $\ge 0$.


## Growth rates

From data sent to me from A.R. Ives related to this paper:

Meisner, M. H., J. P. Harmon, and A. R. Ives. 2014. Temperature effects on
long-term population dynamics in a parasitoid–host system.
*Ecological Monographs* __84__:457–476.

For fecundity, juvenile survival, and adult survival, the dataset had low and high
estimates.
I created two Leslie matrices, one with all of the high estimates and another
with all the lower estimates.
For each of these Leslie matrices, I estimated the intrinsic daily rate of increase
($r$) in aphid populations in the lab as $r = \log(\lambda)$, where $\lambda$ is the
dominant eigenvector of the matrix.
Thus I have fast and slow estimates of population growth, and from these I estimated
the mean and standard deviation of the population-growth parameter.
If the clonal lines from Meisner et al. (2014) are representative of the lines
we currently have, and if these Leslie matrices truly represent the edges of possible
$r$ values, then this standard deviation should be an upper estimate.
The SE of $r$ is what I'll use for the standard deviation of the mean.


```{r load_leslies}
fast <- read_csv("data-raw/leslie_fast.csv", col_names = FALSE,
                 col_types = cols(.default = col_double())) %>% 
    as.matrix() %>% 
    eigen() %>% 
    .[["values"]] %>% 
    .[1] %>% 
    Re() %>% 
    log()
slow <- read_csv("data-raw/leslie_slow.csv", col_names = FALSE,
                 col_types = cols(.default = col_double())) %>% 
    as.matrix() %>% 
    eigen() %>% 
    .[["values"]] %>% 
    .[1] %>% 
    Re() %>% 
    log()
fast; slow
mean(c(fast, slow)); sd(c(fast, slow))
sd(c(fast, slow)) / sqrt(2)
```


The cauchy distribution with scale = 0.005 seems about right for sampling the
among-line SD.

```{r R0_sd_plot}
compare_priors(sd(c(fast, slow)) / sqrt(2), 0.005, xlim = c(0, 0.05)) +
    geom_vline(xintercept = sd(c(fast, slow)) / sqrt(2))
```


## Density dependence

### Among-line distribution

Because Meisner et al. (2014) used cages of 8 plants, I'm setting my mean for $\alpha$
by multiplying their density-dependence estimate ($0.000467$) by 8:
$\bar\alpha = 0.000467 \times 8 = 0.003736$.

They parameterized their density dependence slightly differently from the model I'm
using, so I'm setting the standard deviation for $\alpha$ to be fairly wide.
So for the distribution of mean alpha among all lines, the normal curve below looks
good to me.


```{r alpha_overall_mean_plot}
compare_priors(0.003736, 0.0015, xlim = c(0, 0.01)) +
    geom_vline(xintercept = 0.003736)
```


For sampling mean alphas for each line, I think a scale of 0.0015 is still fine, but
I'll use the cauchy distribution because of its heavier tails.

```{r alpha_line_means_plot}
compare_priors(0.003736, 0.0015, xlim = c(0, 0.01)) +
    geom_vline(xintercept = 0.003736)
```


For the among-line SD distribution, I used a cauchy distribution with location = 0.0015
and scale = 0.0005 because it looks most reasonable to me.

```{r alpha_among_line_sd_plot}
compare_priors(0.0015, 0.0005, xlim = c(0, 0.005)) +
    geom_vline(xintercept = 0.0015)
```


### Within-line distribution

I'm really not sure about the degree to which different plants will affect the
density dependence within an aphid line.
I'll use some raw data from Meisner et al. (2014) to get some idea.
Below is from their cage experiments when they added parasitoid wasps after aphid
densities were allowed to peak (35 days).
I didn't use these estimates for mean alpha because of the potential effect of wasps,
but this is probably as good an estimate of variance among plants that I'll get.
Notice that I'm dividing the max number of aphids by 8 because there were 8 plants
in this experiment.


```{r Meisner_cage_data}
read_csv("data-raw/MeisnerLongtermCageData.csv",
         col_types = "iiiiii") %>% 
    filter(treatment == 0, year == 2009) %>% 
    group_by(replicate) %>% 
    summarize(alpha = 1 / max(aphids / 8)) %>% 
    select(alpha) %>% 
    summarize(sd = sd(alpha)) %>% 
    unlist() %>% 
    signif(3)
```


I'm choosing a cauchy distribution with location of 0.00142 and scale of 0.02 
because I'm pretty uncertain of this parameter.

```{r alpha_within_line_sd_plot}
compare_priors(0.00142, 0.02, xlim = c(0, 0.1)) +
    geom_vline(xintercept = 0.00142)
```



## Process error

I'm not really sure what this should be, but based on simulations, a value of 1 would
be very high.
Thus I chose a cauchy distribution with 0.1 as the location and scale because
this appears about right.

```{r process_error_plot}
compare_priors(0.1, 0.1, xlim = c(0, 1)) +
    geom_vline(xintercept = 0.1)
```

