---
title: "Estimating aphid population parameters"
author: "Lucas A. Nell"
date: "2018-07-03"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Estimating aphid population parameters}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "svglite", 
  fig.width = 6,
  fig.height = 4
)
knitr::opts_knit$set(root.dir = normalizePath(".."))
options(tibble.print_min = 4L, tibble.print_max = 8L)
```
```{r source_Rprofile, echo = FALSE}
source(".Rprofile")
```

```{r library_pkgs, warning=FALSE, echo=FALSE}
suppressPackageStartupMessages({
    library(clonewars)
})
bayesplot::color_scheme_set("viridis")
```


This file will eventually contain the analyses to estimate aphid population-growth
parameters.
It is not yet finished, and much of the code below is not run during the making of
this document.


## Imputing missing data

From [Moritz et al. (2015)](https://arxiv.org/abs/1510.03924), the
seasonal kalman filter from the `zoo` package works well, so I'm creating this function
here to impute missing values below.
We're only missing 1 or 2 time points across a time series of >15, so this
shouldn't have too big an impact on our results.


```{r impute_NAs}
impute <- function(X) {
    if (any(is.na(X))) {
        X_ <- X %>% 
            # Change to weekly frequency for the fit
            zoo::zooreg(frequency = 7) %>%
            zoo::na.StructTS() %>%
            as.numeric()
        return(X_)
    }
    return(X)
}
```


## Load data

```{r load_data}
growth <- 
    load_data(impute_fxn = impute) %>%
    mutate(line = paste(line)) %>% 
    bind_rows(clonewars:::load_pz_data(impute_fxn = impute)) %>% 
    mutate_at(vars(line, rep), funs(factor)) %>% 
    # Now filter out early part of each time series, before N > 6
    # N <= 6 is when the stochasticity associated with only starting with 2 adults
    # appears to be strongest
    group_by(line, rep) %>% 
    filter(1:n() >= which(N > 6)[1]) %>%
    mutate(date = date - min(date)) %>%
    ungroup()
```

This is how many reps I have of each line:

```{r view_reps, echo=FALSE}
growth %>% 
    group_by(line) %>% 
    summarize(n = rep %>% unique() %>% length()) %>% 
    knitr::kable()
```




## See time series data

```{r filter_data}
growth %>% 
    ggplot(aes(date, X, color = factor(rep))) +
    geom_line(size = 0.75) +
    # geom_point() +
    facet_wrap(~ line, nrow = 2) +
    scale_color_brewer(palette = "Dark2", guide = FALSE) +
    NULL
```






## Convert data for stan

```{r convert_data}
model_data <- growth %>% 
    line_data()
str(model_data, give.attr = FALSE)
```


## Run model

```{r run_model, eval = FALSE}
# # Takes ~ 7 min at adapt_delta = 0.9
# set.seed(65136)
# stan_fit <- model_data %>%
#     fit_lines(
#         control = list(adapt_delta = 0.9)
#     )
# write_rds(stan_fit, "data-raw/stan_fit.rds")
stan_fit <- read_rds("data-raw/stan_fit.rds")
```


## Examine model output

```{r examine_model, eval = FALSE}

# http://mc-stan.org/bayesplot/articles/visual-mcmc-diagnostics.html

posterior <- as.array(stan_fit)

mcmc_intervals(posterior, pars = sprintf("R[%i]", 1:8))
mcmc_intervals(posterior, pars = sprintf("A[%i]", 1:8))


apply(rstan::extract(stan_fit, "R", permuted = FALSE), 3, mean)
apply(rstan::extract(stan_fit, "A", permuted = FALSE), 3, mean)

apply(rstan::extract(stan_fit, "s_epsilon", permuted = FALSE), 3, mean)
apply(rstan::extract(stan_fit, "s_epsilon", permuted = FALSE), 3, quantile,
      probs = c(0.025, 0.975))



print(stan_fit, digits = 6, pars = c("R", "A", "s_epsilon"), probs = c())

# parameters of interest
pars_interest <- c(sprintf("A[%i]", 1:8), sprintf("R[%i]", 1:8),
                   paste0(rep(c("", "s_"),each=2), c("theta", "phi")),
                   "hats_phi", "s_epsilon")

# Rhats
rhats <- rhat(stan_fit, pars = pars_interest)
mcmc_rhat(rhats)  + yaxis_text(hjust = 1)

# Neff / N
neff_ratios <- neff_ratio(stan_fit, pars = pars_interest)
mcmc_neff(neff_ratios, size = 2) + yaxis_text(hjust = 1)


# Autocorrelation
mcmc_acf(posterior, pars = pars_interest[1:8], lags = 10)
mcmc_acf(posterior, pars = pars_interest[9:16], lags = 10)
mcmc_acf(posterior, pars = pars_interest[17:length(pars_interest)], lags = 10)


lp <- log_posterior(stan_fit)
np <- nuts_params(stan_fit)

mcmc_trace(posterior, pars = pars_interest[1:8], np = np)
mcmc_trace(posterior, pars = pars_interest[9:16], np = np)
mcmc_trace(posterior, pars = pars_interest[17:length(pars_interest)], np = np)

mcmc_nuts_divergence(np, lp)

mcmc_parcoord(posterior, np = np)

mcmc_scatter(
  posterior,
  pars = c("s_epsilon", "R[6]"),
  transform = list(`R[6]` = "log"),
  np = np
)


growth %>% 
    group_by(line) %>% 
    summarize(n = length(unique(rep))) %>% 
    mutate(line = as.integer(line)) %>% 
    as.data.frame()


pred_df <- make_pred_df(stan_fit, growth)

pred_df %>%
    # filter(rep == 1) %>% 
    # filter(rep == 4, line == "Clover-2017-2") %>%
    ggplot(aes(date, X)) +
    geom_ribbon(aes(ymin = X_lower, ymax = X_upper),
                fill = "gray70", alpha = 0.5) +
    geom_line(color = "firebrick", alpha = 0.5) +
    geom_point(shape = 1, color = "firebrick", alpha = 0.5) +
    geom_line(aes(y = X_pred), color = "dodgerblue3", size = 0.5) +
    # facet_wrap(~ rep + line, nrow = 2) +
    facet_wrap(~ rep + line, nrow = 4) +
    NULL



```


