---
title: "gameofclones internals"
author: "Lucas A. Nell"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{gameofclones internals}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```


This vignette outlines the internal data present in `gameofclones`. 
These data inform some of the default parameters in
`sim_experiments` but are not always identical because `sim_experiments` was
designed to simulate the experimental cages that differed in size
from those used to calculate these parameter values.


The code was initially set up for 2 clones that have different life histories.
The life histories are taken from lab experiments and correspond to
demographic parameters at 20º and 27º C. See 
[Meisner et al. (2014)](http://doi.wiley.com/10.1890/13-1933.1).
They also include parameters related to parasitoid wasps and aphid
resistance to them.


Parameters from this document are saved into 4 lists present as exported 
data for this package:

|               |                                                                    |
|:--------------|:-------------------------------------------------------------------|
| `dev_times`   | Development times for aphids and wasps                             |
| `populations` | Population rates and starting values for aphids and wasps          |
| `wasp_attack` | Wasp attack rate parameters                                        |
| `environ`     | Parameters associated with environmental effects and stochasticity |


Parameters were wrapped into lists to avoid namespace conflicts and to allow me
to use the same names as those in the paper.


# Development times

* `instar_days`: number of days per instar, for low (20º C, `lowT`) and 
  high (27º C, `highT`) temperatures
* `mum_days`: number of days per stage of an aphid being parasitized:
  living and dead ("mummy"), respectively.


```{r instar_mum_days}
dev_times <- list(
    instar_days = list(lowT = c(2, 2, 2, 2, 19), 
                       highT = c(1, 1, 1, 2, 23)),
    mum_days = cbind(7, 3)
)
```



# Populations

These are the stage-structured parameters that specify demographic rates 
by aphid age, for low-growth (`low`) and high-growth (`high`) clonal lines.

* `surv_juv`: juvenile (instar 1-4) daily survival; same for all juvenile ages.
* `surv_adult`: adult survival; different by day. Data are from two clones, 
  one with a high population growth rate, and another with a low rate.
* `repro`: daily fecundity for adults of the two clones, one with a high 
  population growth rate, and another with a low rate.


```{r population1}
populations <- list(
    surv_juv = list(low = 0.9745, high = 0.9849),
    surv_adult = list(
        low = rbind(c(1.0000, 0.9949, 0.9818, 0.9534, 0.8805, 0.8367, 0.8532, 0.8786, 
                      0.8823, 0.8748, 0.8636, 0.8394, 0.8118, 0.8096, 0.8240, 0.8333, 
                      0.7544, 0.5859, 0.4155, 0.2216, rep(0, 180))),
        high = rbind(c(1.0000, 0.9986, 0.9951, 0.9874, 0.9675, 0.9552, 0.9550, 0.9549, 
                       0.9462, 0.8992, 0.8571, 0.8408, 0.8281, 0.8062, 0.7699, 0.7500, 
                       0.7559, 0.7649, 0.7240, 0.4367, rep(0, 180)))),
    repro = list(
        low = rbind(c(0, 2.5925, 4.4312, 5.1403, 5.5190, 5.6633, 5.6010, 5.4577, 5.2904, 
                      5.0613, 4.6970, 3.3577, 1.5946, 1.0817, 0.9666, 0.8333, 0.4689, 
                      0.0709, 0, 0, 0, 0, rep(0, 178))),
        high = rbind(c(0, 3.1975, 5.4563, 6.2996, 6.7372, 6.9030, 6.8210, 6.6100, 
                       6.1962, 5.1653, 4.1837, 3.6029, 3.1023, 2.4799, 1.6909, 1.1750, 
                       1.0148, 0.9096, 0.7821, 0.6430, 0.5000, 0.3531, rep(0, 178))))
)
```


## Other parameters:

These are other population-level parameters for both aphids and wasps.
Parameters `K`, `K_y`, and `s_y` are directly from the paper; `sex_ratio` was 
"hard-coded" into the paper's model but is allowed to change here.
The rest were chosen as reasonable starting values for experiments.


* `K`: aphid density dependence
* `K_y`: parasitized aphid density dependence
* `s_y`: parasitoid adult daily survival
* `sex_ratio`: proportion of female wasps
* `aphids_0`: initial density of aphids
* `wasps_0`: initial densities of wasps
* `prop_resist`: proportion of resistant clones


```{r population2}
populations[["K"]] <- 4.67e-4
populations[["K_y"]] <- 7.33e-4
populations[["s_y"]] <- 0.69
populations[["sex_ratio"]] <- 0.5
populations[["aphids_0"]] <- 20
populations[["wasps_0"]] <- 1
populations[["prop_resist"]] <- 0.05
```




# Wasp attack

These parameters are from equation 6 in Meisner et al. (2014).
Relative attack rates on the different instars are from 
[Ives et al (1999)](http://www.journals.uchicago.edu/doi/10.1086/303269).
The survivals of attacked, resistant aphids is from unpublished code by
Anthony Ives.
This will obviously differ significantly among clones, but was considered a
good starting point for a resistant line.

* `a`: parasitoid attack rate
* `k`: aggregation parameter of the negative binomial distribution
* `h`: parasitoid attack rate handling time
* `rel_attack`: relative attack rates on the different instars
* `attack_surv`: the survivals of singly attacked and multiply attacked
  resistant aphids

```{r attack_akh_rel}
wasp_attack <- list(
    a = 2.32,
    k = 0.35,
    h = 0.008,
    rel_attack = c(0.12, 0.27, 0.39, 0.16, 0.06),
    attack_surv = c(0.9, 0.6)
)
```





# Environment

Parameters associated with environmental effects.
It's assumed that alates (winged aphids) are the primary mode of disperal 
among patches, and alates' wings aren't fully developed until adulthood.
So, the day at which aphids can begin dispersing unsurprisingly depends on 
how quickly the aphids are developing.

* `harvest_surv`: aphid survival rate at harvesting
* `disp_aphid`: dispersal rates between fields for aphids, adult wasps
* `disp_wasp`: dispersal rates between fields for aphids, adult wasps
* `pred_rate`: predation rate for aphids and non-adult wasps
* `cycle_length`: time between harvests (typical for alfalfa)
* `field_disp_start`: when aphids disperse by flying, for low (20º C; `lowT`) and 
  high (27º C; `highT`) temperatures


```{r environ}
environ <- list(
    harvest_surv = 0.05,
    disp_aphid = 0.05,
    disp_wasp = 1,
    pred_rate = 0.8,
    cycle_length = 30,
    field_disp_start = list(
        lowT = sum(dev_times$instar_days$lowT[1:4]) + 1, 
        highT = sum(dev_times$instar_days$highT[1:4]) + 1)
)
```



Parameters associated with environmental stochasticity.

`rho` was estimated at 1.0 in Meisner et al. (2014), but that estimate
"... was largely an artifact of the fitting procedure" (p 468).
The number here is from unpublished code by Anthony Ives that works to
simulate the cages we are using for experiments.

For their simulations, Meisner et al. (2014) multiplied the original estimated 
values of `sigma_x` and `sigma_y` by half to "emphasize the demographic 
stochasticity" (p 469).
The values below are the original estimated values, not those used for their 
simulations.

* `sigma_x`: environmental std dev for aphids
* `sigma_y`: environmental std dev for wasps
* `rho`: environmental correlation among instars

```{r process_error}
environ[["sigma_x"]] <- 0.44
environ[["sigma_y"]] <- 0.70
environ[["rho"]] <- signif(2/(1+exp(-environ[["sigma_y"]]))-1, 2)
```




# Saving data

Saving the created objects as internal data for `gameofclones`.

```{r save_data, eval = FALSE}
usethis::use_data(dev_times, populations, wasp_attack, environ)
```
