---
title: "gameofclones parameters"
author: "Lucas A. Nell"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{gameofclones parameters}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
library(gameofclones)
knitr::opts_chunk$set(comment = "")
```


This vignette describes the arguments that can be changed for a given aphid line.
It also includes arguments on parasitoid wasps and how aphids interact with wasps, 
but those aren't (yet) included in the simulations.
They may be added later.


# Argument descriptions

- `aphid_density_0`: Starting aphid density.
- `aphid_surv_juv`: Aphid juvenile survival.
- `aphid_surv_adult`: Vector of aphid adult survivals by stage.
- `aphid_repro`: Vector of aphid reproductive rates by stage.
- `K`: Aphid density dependence.
- `wasp_density_0`: Starting wasp density.
- `K_y`: Wasp density dependence.
- `s_y`: Wasp adult daily survival.
- `wasp_sex_ratio`: Proportion of female wasps.
- `aphid_instar_days`: Vector of the number of stages (days) per aphid instar.
- `mum_days`: Vector for the number of days per parasitized-aphid stage. It should be 
  of length 2, representing stages where the parasitized aphid is alive and dead, 
  respectively.
- `rel_attack`: Vector of relative wasp attack rates by aphid stage.
- `a`: Overall parasitoid attack rate.
- `k`: Aggregation parameter of the nbinom distribution.
- `h`: Parasitoid attack rate handling time.
- `attack_surv`: Vector of length 2 with survival rates of singly & multiply attacked 
  aphids, respectively.
- `sigma_x`: Environmental standard deviation for aphids.
- `sigma_y`: Environmental standard deviation for parasitoids.
- `rho`: Environmental correlation among instars.
- `demog_mult`: Multiplier for demographic stochasticity.
- `harvest_surv`: Survival rate for living aphids during a harvest.
- `disp_aphid`: Dispersal rate for aphids.
- `disp_wasp`: Dispersal rate for wasps.
- `disp_start`: Stage in which dispersal starts in aphids.
- `pred_rate`: Predation on aphids and mummies.



# Default values

Below, the "Default-value expression" is the code necessary to access and, in some 
cases, manipulate an internal-data object to compute the default value.
Here, `NA` means it is not stored internally.
Note also that although the default value for `attack_surv`  is not stored internally,
the object `gameofclones::wasp_attack$attack_surv` stores values for a resistant line of aphids.

I first created this function for the calls to `Reduce`.
```{r meanfun}
mf <- function(x, y) (x + y) / 2
```


| Argument name       | Default-value expression                  | Default value                                            |
|:--------------------|:------------------------------------------|:---------------------------------------------------------|
| `aphid_density_0`   | `gameofclones::populations$aphids_0`               | <code>`r gameofclones::populations$aphids_0`</code>               |
| `aphid_surv_juv`    | `Reduce(mf,gameofclones::populations$surv_juv)`    | <code>`r Reduce(mf,gameofclones::populations$surv_juv)`</code>    |
| `aphid_surv_adult`  | `Reduce(mf,gameofclones::populations$surv_adult)`  | <code>`r Reduce(mf,gameofclones::populations$surv_adult)`</code>  |
| `aphid_repro`       | `Reduce(mf,gameofclones::populations$repro)`       | <code>`r Reduce(mf,gameofclones::populations$repro)`</code>       |
| `K`                 | `gameofclones::populations$K`                      | <code>`r gameofclones::populations$K`</code>                      |
| `wasp_density_0`    | `gameofclones::populations$wasps_0`                | <code>`r gameofclones::populations$wasps_0`</code>                |
| `K_y`               | `gameofclones::populations$K_y`                    | <code>`r gameofclones::populations$K_y`</code>                    |
| `s_y`               | `gameofclones::populations$s_y`                    | <code>`r gameofclones::populations$s_y`</code>                    |
| `wasp_sex_ratio`    | `gameofclones::populations$sex_ratio`              | <code>`r gameofclones::populations$sex_ratio`</code>              |
| `aphid_instar_days` | `gameofclones::dev_times$instar_days$highT`        | <code>`r gameofclones::dev_times$instar_days$highT`</code>        |
| `mum_days`          | `gameofclones::dev_times$mum_days`                 | <code>`r gameofclones::dev_times$mum_days`</code>                 |
| `rel_attack`        | `gameofclones::wasp_attack$rel_attack$highT`       | <code>`r gameofclones::wasp_attack$rel_attack$highT`</code>       |
| `a`                 | `gameofclones::wasp_attack$a`                      | <code>`r gameofclones::wasp_attack$a`</code>                      |
| `k`                 | `gameofclones::wasp_attack$k`                      | <code>`r gameofclones::wasp_attack$k`</code>                      |
| `h`                 | `gameofclones::wasp_attack$h`                      | <code>`r gameofclones::wasp_attack$h`</code>                      |
| `attack_surv`       | `gameofclones::wasp_attack$attack_surv / 2`        | <code>`r gameofclones::wasp_attack$attack_surv / 2`</code>        |
| `sigma_x`           | `gameofclones::environ$sigma_x`                    | <code>`r gameofclones::environ$sigma_x`</code>                    |
| `sigma_y`           | `gameofclones::environ$sigma_y`                    | <code>`r gameofclones::environ$sigma_y`</code>                    |
| `rho`               | `gameofclones::environ$rho`                        | <code>`r gameofclones::environ$rho`</code>                        |
| `demog_mult`        | `NA`                                      | <code>`r 1`</code>                                       |
| `harvest_surv`      | `gameofclones::environ$harvest_surv`               | <code>`r gameofclones::environ$harvest_surv`</code>               |
| `disp_aphid`        | `gameofclones::environ$disp_aphid`                 | <code>`r gameofclones::environ$disp_aphid`</code>                 |
| `disp_wasp`         | `gameofclones::environ$disp_wasp`                  | <code>`r gameofclones::environ$disp_wasp`</code>                  |
| `disp_start`        | `gameofclones::environ$disp_start$highT`           | <code>`r gameofclones::environ$disp_start$highT`</code>           |
| `pred_rate`         | `gameofclones::environ$pred_rate`                  | <code>`r gameofclones::environ$pred_rate`</code>                  |




# High tunnel simulations

For the simulations Tony sent that he'd designed to use for the high tunnels, 
he used two lines, one susceptible and one resistant to parasitoids, and no 
process error.
The days at which harvests occurred were handled not using a cycle length and 
offset, but with a vector of harvest times; this made little difference, though.
The following parameter values were different from default values:

```{r dens_vals, echo = FALSE}
sus_dens <- (1 - gameofclones::populations$prop_resist) * gameofclones::populations$aphids_0
res_dens <- gameofclones::populations$prop_resist * gameofclones::populations$aphids_0

res_surv <- gameofclones::wasp_attack$attack_surv

sus_adult <- gameofclones::populations$surv_adult$high[1:32]
res_adult <- gameofclones::populations$surv_adult$low[1:32]

sus_repro <- gameofclones::populations$repro$high[1:32]
res_repro <- gameofclones::populations$repro$low[1:32]
```

| Argument name      | Susceptible                | Resistant                  |
|:-------------------|:---------------------------|:---------------------------|
| `aphid_density_0`  | <code>`r sus_dens`</code>  | <code>`r res_dens`</code>  |
| `aphid_surv_adult` | <code>`r sus_adult`</code> | <code>`r res_adult`</code> |
| `aphid_repro`      | <code>`r sus_repro`</code> | <code>`r res_repro`</code> |
| `K`                | `0.0005`                   | `0.0005`                   |
| `K_y`              | `0.0006`                   | `0.0006`                   |
| `s_y`              | `0.55`                     | `0.55`                     |
| `a`                | `2.5`                      | `2.5`                      |
| `k`                | `0.1811`                   | `0.1811`                   |
| `h`                | `0.0363`                   | `0.0363`                   |
| `attack_surv`      | <code>`r c(0,0)`</code>    | <code>`r res_surv`</code>  |
| `sigma_x`          | `0`                        | `0`                        |
| `sigma_y`          | `0`                        | `0`                        |
| `rho`              | `0`                        | `0`                        |
| `demog_mult`       | `0`                        | `0`                        |

