---
title: "Simulating eco-evo experiments"
author: "Lucas A. Nell"
date: "2/1/2021"
output: pdf_document
editor_options: 
  chunk_output_type: console
urlcolor: blue
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, dev = "cairo_pdf", echo = FALSE, 
                      fig.width = 5, fig.height = 3)
suppressPackageStartupMessages({
    library(tidyverse)
    library(clonewars)
})
```


These simulations are to see if doing eco-evo experiments will actually be
interesting if one of the lines (UT3) is completely resistant to parasitism.
I started with a model from a 
[previous paper](http://doi.wiley.com/10.1890/13-1933.1)
from Tony and others, then I edited it to conform to our conditions
(most notably, it explicitly simulates alate production and plant death).

Note that figures are at the end of the document.

I'm simulating 100 days of interacting aphid and parasitoid wasp populations.
There are 4 patches of plants, and each patch contains 4 plants.
Every 25 days, we replace a plant from each patch, which causes mortality
for any aphids or mummies on that plant.
(Mummies are aphids that have been killed by being parasitized, and are
about to have adult wasps emerge from their bodies.)
Movement between patches must happen via alates (winged aphids), and 
rates of alate production increase with the number of aphids on the patch.
There are two aphid lines:
The susceptible line is susceptible to parasitism but has a higher growth rate.
The resistant line is resistant to parasitism but has a lower growth rate
(attack by wasps induces some mortality, though).
Shifts in relative frequencies of susceptible and resistant lines
represents evolution in the aphid population.




```{r args-setup}
# This is constant:
wia5d <- clonal_line("susceptible",
                     density_0 = matrix(c(rep(0, 3), 16, rep(0, 6)), 5, 2),
                     surv_juv_apterous = "high",
                     surv_adult_apterous = "high",
                     repro_apterous = "high")


# The call to `sim_clonewars` can vary by the UT3 line plus
# wasp mortality (`s_y`)
#   * `.res` should be FALSE, TRUE, or c(1, 1)
#   * `.sjp` should be "low" or 0
#   * `.s_y` should probably be a multiple of `populations$s_y`
do_sims <- function(.res, .sjp, .s_y, ...) {
    .ut3 <- clonal_line("resistant",
                        density_0 = matrix(c(rep(0, 3), 16, rep(0, 6)), 5, 2),
                        resistant = .res,
                        surv_juv_paras = .sjp,
                        surv_juv_apterous = "low",
                        surv_adult_apterous = "low",
                        repro_apterous = "low")
    sims <- sim_clonewars(n_reps = 1, 
                          clonal_lines = c(.ut3, wia5d),
                          s_y = .s_y,
                          alate_b0 = -4, alate_b1 = 8/2000,
                          max_N = 0, max_plant_age = 25,
                          clear_surv = 0.5,
                          mean_K = formals(sim_clonewars)$mean_K * 4,
                          sd_K = formals(sim_clonewars)$sd_K * 4,
                          no_error = TRUE,
                          sex_ratio = 0.66,
                          wasp_delay = 5,
                          wasp_density_0 = 4, 
                          ...)
    sims[["mummies"]] <- sims %>%
                  .[["aphids"]] %>%
                  filter(type == "mummy") %>%
                  group_by(rep, time) %>%
                  summarize(N = sum(N), .groups = "drop")
    sims[["aphids"]] <- sims %>%
        .[["aphids"]] %>%
        filter(type != "mummy") %>%
        group_by(rep, time, line) %>%
        summarize(N = sum(N), .groups = "drop")
    
    return(sims)
}

# For plotting:
plot_sims <- function(.sims) {
    
    mod <- max(max(.sims$wasps$wasps), max(.sims$mummies$N)) / 
        max(.sims$aphids$N)
    
    .sims %>%
        .[["aphids"]] %>%
        mutate(line = factor(line, levels = c("resistant", "susceptible"))) %>%
        ggplot(aes(time, N)) +
        geom_area(data = .sims %>%
                      .[["mummies"]] %>%
                      mutate(N = N / mod),
                  fill = "gray70", color = NA) +
        geom_area(data = .sims %>%
                      .[["wasps"]] %>%
                      mutate(N = wasps / mod),
                  fill = "violetred", alpha = 0.5, color = NA) +
        geom_line(aes(color = line), size = 1) +
        scale_color_manual(NULL, values = c("chartreuse3", "firebrick")) +
        scale_linetype_manual(values = c(1, 2)) +
        scale_y_continuous("Aphid abundance",
                           sec.axis = sec_axis(~ . * mod,
                                               "Wasp & mummy abundance"))
}

fig1_cap <- paste("Aphids, mummies, and wasps over time. Gray shaded areas",
                  "are mummies, pink shaded areas are adult wasps,",
                  "and lines are aphids  (green for resistant, red for",
                  "susceptible).")
fig2_cap <- paste("Aphids, mummies, and wasps over time, with varying adult",
                  "wasp survival.",
                  "Sub-panels separate the adult wasp survival rates used,",
                  "and labels above sub-panels indicate how they relate",
                  "to the default value ($\\hat{s}_y$).",
                  "Gray shaded areas",
                  "are mummies, pink shaded areas are adult wasps,",
                  "and lines are aphids (green for resistant, red for",
                  "susceptible).")
```


With everything at their default values (other than those mentioned above),
we get the results shown in Figure 1.
The resistant line excludes the susceptible line and wasps go extinct
after a brief period of substantial parasitism.
This is the type of uninteresting results I was worried about.


Past experiments have shown that wasps can decimate aphid populations in 
closed systems.
A potential solution to this problem is to reduce the amount of time
that adults can spend parasitizing aphids.
In the model, this can be implemented by reducing adult wasp survival.
When we divide the default adult survival ($\hat{s}_y$) by some different 
values, we get the results shown in Figure 2.


This is much more interesting!
In all cases, both lines coexist over the 100 days, but
the resistant line becomes less dominant the more we decrease adult wasp
survival.
The nice thing about adult wasp survival is that it's easy to control,
so this is quite encouraging.
These results are also robust to wasp attack not causing the resistant line
any mortality (not shown, but you can trust me).




```{r og-sims, fig.cap=fig1_cap}
#   * `.res` should be FALSE, TRUE, or c(1, 1)
#   * `.sjp` should be "low" or 0
#   * `.s_y` should probably be a multiple of `populations$s_y`
sims <- do_sims(.res = TRUE, .sjp = 0, .s_y = populations$s_y)


p <- plot_sims(sims) +
    geom_text(data = tibble(line = factor(c("resistant", "susceptible"), 
                                          levels = c("resistant", "susceptible")),
                            time = 100, N = c(6000, 500)), 
              aes(label = line, color = line), size = 10 / 2.8, hjust = 1) +
    geom_text(data = tibble(line = "mummies", time = 22, N = 6000), 
              aes(label = line), color = "gray50", size = 10 / 2.8, hjust = 1) +
    geom_text(data = tibble(line = "wasps", time = 48, N = 1000),
              aes(label = line), color = "violetred", size = 10 / 2.8, hjust = 0) +
    theme(legend.position = "none")

p

# ggsave("~/Desktop/test.pdf", p, width = 5, height = 3)

```



```{r change-s_y, fig.height = 8, fig.cap=fig2_cap}

sims <- map(c(4, 16, 64), function(.x) {
    s <- do_sims(.res = TRUE, .sjp = 0, .s_y = populations$s_y / .x)
    s[["aphids"]] <- mutate(s[["aphids"]], s_y = sprintf("hat(s)[y] / %i", .x))
    s[["mummies"]] <- mutate(s[["mummies"]], s_y = sprintf("hat(s)[y] / %i", .x))
    s[["wasps"]] <- mutate(s[["wasps"]], s_y = sprintf("hat(s)[y] / %i", .x))
    return(s)
    })
aphids <- map_dfr(sims, ~ .x[["aphids"]]) %>% 
    select(-rep) %>% 
    mutate(s_y = factor(s_y, levels = sprintf("hat(s)[y] / %i", c(4, 16, 64))))
mummies <- map_dfr(sims, ~ .x[["mummies"]]) %>% 
    select(-rep) %>% 
    mutate(s_y = factor(s_y, levels = sprintf("hat(s)[y] / %i", c(4, 16, 64))))
wasps <- map_dfr(sims, ~ .x[["wasps"]]) %>% 
    select(-rep) %>% 
    mutate(s_y = factor(s_y, levels = sprintf("hat(s)[y] / %i", c(4, 16, 64))))


mod <- max(max(wasps$wasps), max(mummies$N)) / max(aphids$N)

p <- aphids %>%
    mutate(line = factor(line, levels = c("resistant", "susceptible"))) %>%
    ggplot(aes(time, N)) +
    geom_area(data = mummies %>%
                  mutate(N = N / mod),
              fill = "gray70", color = NA) +
    geom_area(data = wasps %>%
                  mutate(N = wasps / mod),
              fill = "violetred", alpha = 0.5, color = NA) +
    geom_line(aes(color = line), size = 1) +
    facet_wrap(~ s_y, labeller = label_parsed, ncol = 1) +
    scale_color_manual(NULL, values = c("chartreuse3", "firebrick")) +
    scale_linetype_manual(values = c(1, 2)) +
    scale_y_continuous("Aphid abundance",
                       sec.axis = sec_axis(~ . * mod,
                                           "Wasp & mummy abundance")) +
    theme(legend.position = "none")

p

# ggsave("~/Desktop/test.pdf", p, width = 5, height = 8)


```




```{r no-mort, eval=FALSE}

do_sims(.res = TRUE, .sjp = 0, .s_y = populations$s_y / 16) %>% 
    plot_sims()

```

