---
title: "Simulating eco-evo experiments"
author: "Lucas A. Nell"
date: "2/1/2021"
output: pdf_document
editor_options: 
  chunk_output_type: console
urlcolor: blue
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, dev = "cairo_pdf", echo = FALSE, 
                      fig.width = 5, fig.height = 3)
knitr::opts_knit$set(root.dir = "../")
suppressPackageStartupMessages({
    library(tidyverse)
    library(clonewars)
})
```


These simulations are to see if doing eco-evo experiments will actually be
interesting if one of the lines (UT3) is completely resistant to parasitism.
I started with a model from a 
[previous paper](http://doi.wiley.com/10.1890/13-1933.1)
from Tony and others, then I edited it to conform to our conditions
(most notably, it explicitly simulates alate production and plant death).

Note that figures are at the end of the document.

I'm simulating 100 days of interacting aphid and parasitoid wasp populations.
There are 4 patches of plants, and each patch contains 4 plants.
Every 25 days, we replace a plant from each patch, which causes mortality
for any aphids or mummies on that plant.
(Mummies are aphids that have been killed by being parasitized, and are
about to have adult wasps emerge from their bodies.)
Movement between patches must happen via alates (winged aphids), and 
rates of alate production increase with the number of aphids on the patch.
There are two aphid lines:
The susceptible line is susceptible to parasitism but has a higher growth rate.
The resistant line is resistant to parasitism but has a lower growth rate
(attack by wasps induces some mortality, though).
Shifts in relative frequencies of susceptible and resistant lines
represents evolution in the aphid population.



```{r args-setup}
# This is constant:
wia5d <- clonal_line("susceptible",
                     density_0 = matrix(c(rep(0, 3), 16, rep(0, 6)), 5, 2))  #, 
                     # surv_juv_apterous = "high", 
                     # surv_adult_apterous = "high", 
                     # repro_apterous = "high")


# The call to `sim_clonewars` can vary by the UT3 line plus
# wasp mortality (`s_y`)
#   * `.res` should be FALSE, TRUE, or c(1, 1)
#   * `.sjp` should be "low" or 0
#   * `.s_y` should probably be a multiple of `populations$s_y`
do_sims <- function(.res, .sjp, .s_y, ...) {
    .ut3 <- clonal_line("resistant",
                        density_0 = matrix(c(rep(0, 3), 16, rep(0, 6)), 5, 2),
                        resistant = .res,
                        surv_paras = .sjp,
                        surv_juv_apterous = "low",
                        surv_adult_apterous = "low",
                        repro_apterous = "low")
    sims <- sim_clonewars(n_reps = 1, 
                          clonal_lines = c(.ut3, wia5d),
                          s_y = .s_y,
                          alate_b0 = -4, alate_b1 = 8/2000,
                          max_N = 0, max_plant_age = 25,
                          clear_surv = 0.5,
                          mean_K = formals(sim_clonewars)$mean_K * 4,
                          sd_K = formals(sim_clonewars)$sd_K * 4, 
                          no_error = TRUE,
                          sex_ratio = 0.66,
                          wasp_delay = 0,
                          wasp_density_0 = 4, 
                          ...)
    sims[["mummies"]] <- sims %>%
                  .[["aphids"]] %>%
                  filter(type == "mummy") %>%
                  group_by(rep, time) %>%
                  summarize(N = sum(N), .groups = "drop")
    sims[["aphids"]] <- sims %>%
        .[["aphids"]] %>%
        filter(type != "mummy") %>%
        group_by(rep, time, line) %>%
        summarize(N = sum(N), .groups = "drop")
    
    return(sims)
}

# For plotting:
plot_sims <- function(.sims) {
    
    mod <- max(max(.sims$wasps$wasps), max(.sims$mummies$N)) / 
        max(.sims$aphids$N)
    
    .sims %>%
        .[["aphids"]] %>%
        mutate(line = factor(line, levels = c("resistant", "susceptible"))) %>%
        ggplot(aes(time, N)) +
        geom_area(data = .sims %>%
                      .[["mummies"]] %>%
                      mutate(N = N / mod),
                  fill = "gray70", color = NA) +
        geom_area(data = .sims %>%
                      .[["wasps"]] %>%
                      mutate(N = wasps / mod),
                  fill = "violetred", alpha = 0.5, color = NA) +
        geom_line(aes(color = line), size = 1) +
        scale_color_manual(NULL, values = c("chartreuse3", "firebrick")) +
        scale_linetype_manual(values = c(1, 2)) +
        scale_y_continuous("Aphid abundance",
                           sec.axis = sec_axis(~ . * mod,
                                               "Wasp & mummy abundance"))
}

fig1_cap <- paste("Aphids, mummies, and wasps over time. Gray shaded areas",
                  "are mummies, pink shaded areas are adult wasps,",
                  "and lines are aphids  (green for resistant, red for",
                  "susceptible).")
fig2_cap <- paste("Aphids, mummies, and wasps over time, with varying adult",
                  "wasp survival.",
                  "Sub-panels separate the adult wasp survival rates used,",
                  "and labels above sub-panels indicate how they relate",
                  "to the default value ($\\hat{s}_y$).",
                  "Gray shaded areas",
                  "are mummies, pink shaded areas are adult wasps,",
                  "and lines are aphids (green for resistant, red for",
                  "susceptible).")
```



### Standard simulations

With everything at their default values (other than those mentioned above),
we get the results shown in Figure 1.
The resistant line excludes the susceptible line and wasps go extinct
after a brief period of substantial parasitism.
This is the type of uninteresting results I was worried about.





```{r og-sims, fig.cap=fig1_cap}
#   * `.res` should be FALSE, TRUE, or c(1, 1)
#   * `.sjp` should be "low" or 0
#   * `.s_y` should probably be a multiple of `populations$s_y`
sims <- do_sims(.res = TRUE, .sjp = 0, .s_y = populations$s_y)


p <- plot_sims(sims) +
    geom_text(data = tibble(line = factor(c("resistant", "susceptible"),
                                          levels = c("resistant", "susceptible")),
                            time = 100, N = c(6000, 500)),
              aes(label = line, color = line), size = 9 / 2.8, hjust = 1) +
    geom_text(data = tibble(line = "mummies", time = 50, N = 6000), 
              aes(label = line), color = "gray50", size = 9 / 2.8, hjust = 1) +
    geom_text(data = tibble(line = "wasps", time = 75, N = 2000),
              aes(label = line), color = "violetred", size = 9 / 2.8, hjust = 0) +
    theme(legend.position = "none")

# p

# ggsave("~/Desktop/basic_sims.pdf", p, width = 5, height = 3)

```


### Changing adult wasp survival


Past experiments have shown that wasps can decimate aphid populations in 
closed systems.
A potential solution to this problem is to reduce the amount of time
that adults can spend parasitizing aphids.
In the model, this can be implemented by reducing adult wasp survival.
When we divide the default adult survival ($\hat{s}_y$) by some different 
values, we get the results shown in Figure 2.


This is much more interesting!
In all cases, both lines coexist over the 100 days, but
the resistant line becomes less dominant the more we decrease adult wasp
survival.
The nice thing about adult wasp survival is that it's easy to control,
so this is quite encouraging.
These results are also robust to wasp attack not causing the resistant line
any mortality (not shown, but you can trust me).



```{r change-s_y, fig.height = 8, fig.cap=fig2_cap}

lvls <- c(4, 8, 16)

sy_sims <- map(lvls, function(.x) {
    s <- do_sims(.res = TRUE, .sjp = 0, .s_y = populations$s_y / .x)
    s[["aphids"]] <- mutate(s[["aphids"]], s_y = sprintf("hat(s)[y] / %i", .x))
    s[["mummies"]] <- mutate(s[["mummies"]], s_y = sprintf("hat(s)[y] / %i", .x))
    s[["wasps"]] <- mutate(s[["wasps"]], s_y = sprintf("hat(s)[y] / %i", .x))
    return(s)
    })
sy_aphids <- map_dfr(sy_sims, ~ .x[["aphids"]]) %>% 
    select(-rep) %>% 
    mutate(s_y = factor(s_y, levels = sprintf("hat(s)[y] / %i", lvls)))
sy_mummies <- map_dfr(sy_sims, ~ .x[["mummies"]]) %>% 
    select(-rep) %>% 
    mutate(s_y = factor(s_y, levels = sprintf("hat(s)[y] / %i", lvls)))
sy_wasps <- map_dfr(sy_sims, ~ .x[["wasps"]]) %>% 
    select(-rep) %>% 
    mutate(s_y = factor(s_y, levels = sprintf("hat(s)[y] / %i", lvls)))


sy_mod <- max(max(sy_wasps$wasps), max(sy_mummies$N)) / max(sy_aphids$N)

sy_p <- sy_aphids %>%
    mutate(line = factor(line, levels = c("resistant", "susceptible"))) %>%
    ggplot(aes(time, N)) +
    geom_area(data = sy_mummies %>%
                  mutate(N = N / sy_mod),
              fill = "gray70", color = NA) +
    geom_area(data = sy_wasps %>%
                  mutate(N = wasps / sy_mod),
              fill = "violetred", alpha = 0.5, color = NA) +
    geom_line(aes(color = line), size = 1) +
    facet_wrap(~ s_y, labeller = label_parsed, ncol = 1) +
    scale_color_manual(NULL, values = c("chartreuse3", "firebrick")) +
    scale_linetype_manual(values = c(1, 2)) +
    scale_y_continuous("Aphid abundance",
                       sec.axis = sec_axis(~ . * sy_mod,
                                           "Wasp & mummy abundance")) +
    theme(legend.position = "none")

sy_p

# ggsave("~/Desktop/vary_sy.pdf", sy_p, width = 5, height = 8)


```





### Stability to starting conditions

Below shows that the coexistence results can be relatively stable
to varying starting conditions, especially as adult wasp survival decreases.
In this case, we're varying the proportion of starting aphids that are 
resistant to parasitism ($p_res$).


```{r stability-p_res-sims-caption}
fig3_cap <- paste("Aphids and wasps over time, with varying adult",
                  "wasp survival and starting proportion of aphids with",
                  "resistance ($p_{res}$).",
                  "Sub-panel columns separate the adult wasp survival rates used,",
                  "and labels above indicate how they relate",
                  "to the default value ($\\hat{s}_y$).",
                  "Sub-panel rows separate starting proportions of aphid",
                  "resistance.",
                  "Gray lines are adult wasps, green are resistant aphids,",
                  "and red are susceptible aphids.")
```




## LEFT OFF
<!-- See line that starts `stable_sims <- ...` -->

```{r stability-p_res-sims, fig.height=6.5, fig.width=6.5, fig.cap=fig3_cap}


do_stable_sims <- function(.p_res, .s_y, ...) {
    
    # .p_res <- 0.5; .s_y <- populations$s_y * 0.8
    
    .n <- 2 * wia5d$density_0[wia5d$density_0 > 0]
    # .n <- 500
    
    .wia5d <- wia5d
    .wia5d$density_0[.wia5d$density_0 > 0] <- .n * (1 - .p_res)
    
    .ut3 <- clonal_line("resistant",
                        density_0 = matrix(c(rep(0, 3), .n * .p_res, 
                                             rep(0, 6)), 5, 2),
                        resistant = TRUE,
                        surv_paras = 0,
                        surv_juv_apterous = "low",
                        surv_adult_apterous = "low",
                        repro_apterous = "low")
    sims <- sim_clonewars(n_reps = 1, 
                          clonal_lines = c(.ut3, .wia5d),
                          s_y = .s_y,
                          alate_b0 = -4, alate_b1 = 8/2000,
                          max_N = 0, max_plant_age = 25,
                          clear_surv = 0.5,
                          mean_K = formals(sim_clonewars)$mean_K * 4,
                          sd_K = formals(sim_clonewars)$sd_K * 4,
                          no_error = TRUE,
                          sex_ratio = 0.66,
                          wasp_delay = 0,
                          wasp_density_0 = 4, 
                          ...)
    sims[["mummies"]] <- sims %>%
        .[["aphids"]] %>%
        filter(type == "mummy") %>%
        group_by(rep, time) %>%
        summarize(N = sum(N), .groups = "drop") %>% 
        mutate(s_y = .s_y, p_res = .p_res)
    sims[["aphids"]] <- sims %>%
        .[["aphids"]] %>%
        filter(type != "mummy") %>%
        group_by(rep, time, line) %>%
        summarize(N = sum(N), .groups = "drop") %>% 
        mutate(s_y = .s_y, p_res = .p_res)
    sims[["wasps"]] <- sims %>%
        .[["wasps"]] %>%
        mutate(s_y = .s_y, p_res = .p_res)
    
    return(sims)
}


stable_fct <- function(.x) {
    if ("s_y" %in% colnames(.x)) {
        .x <- .x %>% 
            mutate(s_y = factor(s_y, levels = populations$s_y / stable_lvls, 
                                labels = sprintf("hat(s)[y] / %i", 
                                                 stable_lvls)))
    }
    if ("pert" %in% colnames(.x)) {
        .x <- .x %>% 
            mutate(pert = factor(pert, levels = c("none", "susceptible", 
                                                  "resistant", "wasps")))
    }
    if ("p_res" %in% colnames(.x)) {
        .x <- .x %>% 
            mutate(p_res = factor(p_res, levels = stable_p_res,
                                  labels = sprintf("p[res] == %.1f", 
                                                   stable_p_res)))
    }
    if ("line" %in% colnames(.x)) {
        .x <- .x %>% 
            mutate(line = factor(line, levels = c("resistant", "susceptible")))
    }
    
    return(.x)
}



stable_lvls <- c(8, 16, 32)
stable_p_res <- seq(0.2, 0.8, 0.1)



# Takes ~1 min
stable_sims <- crossing(.p_res = stable_p_res,
                        .s_y = populations$s_y / stable_lvls) %>%
    # pmap(do_stable_sims, max_t = 1000,
    #      h = 0,
    #      rel_attack = wasp_attack$rel_attack$lowT * 0.1)
    pmap(do_stable_sims, max_t = 1000)

# saveRDS(stable_sims, "under_constr/stable_sims.rds")
# 
# stable_sims <- readRDS("under_constr/stable_sims.rds")


stable_aphids <- map_dfr(stable_sims, ~ .x[["aphids"]]) %>%
    select(-rep) %>%
    stable_fct()
stable_mummies <- map_dfr(stable_sims, ~ .x[["mummies"]]) %>%
    select(-rep) %>%
    stable_fct()
stable_wasps <- map_dfr(stable_sims, ~ .x[["wasps"]]) %>%
    select(-rep) %>%
    stable_fct()

stable_mod <- max(stable_wasps$wasps) / max(stable_aphids$N)


stab_p <- stable_aphids %>%
    ggplot(aes(time, N)) +
    geom_line(data = stable_wasps %>%
                  mutate(N = wasps / stable_mod),
              color = "gray70", size = 0.25) +
    geom_line(aes(color = line), size = 0.25) +
    geom_text(data = tibble(line = factor(c("resistant", "susceptible"),
                                          levels = c("resistant", "susceptible")),
                            s_y = populations$s_y / stable_lvls[1],
                            p_res = stable_p_res[1],
                            time = 1000, N = c(1500, 6000)) %>%
                  stable_fct(),
              aes(label = line, color = line), size = 10 / 2.8, hjust = 1, vjust = 0.5) +
    # geom_text(data = tibble(s_y = populations$s_y / stable_lvls[3],
    #                         p_res = stable_p_res[1],
    #                         time = 100, N = 8000) %>%
    #               stable_fct(),
    #           label = "wasps", color = "gray50", size = 10 / 2.8, hjust = 0) +
    facet_grid(p_res ~ s_y, labeller = label_parsed) +
    scale_color_manual(NULL, values = c("chartreuse3", "firebrick")) +
    scale_linetype_manual(values = c(1, 2)) +
    scale_y_continuous("Aphid abundance", #breaks = 0:2 * 4000,
                       sec.axis = sec_axis(~ . * stable_mod,
                                           "Adult wasp abundance")) +
    theme(legend.position = "none")

stab_p
stab_p + coord_cartesian(xlim = c(0, 100))


stable_sims2 <- c(crossing(.p_res = stable_p_res,
                                   .s_y = populations$s_y) %>%
                              pmap(do_stable_sims, max_t = 1000, 
                                   max_mum_density = 10),
                          crossing(.p_res = stable_p_res,
                                   .s_y = populations$s_y / 16) %>%
                              pmap(do_stable_sims, max_t = 1000))


stable_aphids2 <- map_dfr(stable_sims2, ~ .x[["aphids"]]) %>%
    mutate(sy = factor(s_y)) %>% 
    select(-rep, -s_y) %>% 
    stable_fct()
stable_mummies2 <- map_dfr(stable_sims2, ~ .x[["mummies"]]) %>% 
    mutate(sy = factor(s_y)) %>% 
    select(-rep, -s_y) %>% 
    stable_fct()
stable_wasps2 <- map_dfr(stable_sims2, ~ .x[["wasps"]]) %>% 
    mutate(sy = factor(s_y)) %>% 
    select(-rep, -s_y) %>% 
    stable_fct()

stable_mod2 <- max(stable_wasps2$wasps) / max(stable_aphids2$N)

stab_p2 <- stable_aphids2 %>%
    ggplot(aes(time, N)) +
    geom_line(data = stable_wasps2 %>%
                  mutate(N = wasps / stable_mod2),
              color = "gray70", size = 0.25) +
    geom_line(aes(color = line), size = 0.25) +
    facet_grid(p_res ~ sy, labeller = label_parsed) +
    scale_color_manual(NULL, values = c("chartreuse3", "firebrick")) +
    scale_linetype_manual(values = c(1, 2)) +
    scale_y_continuous("Aphid abundance", #breaks = 0:2 * 4000,
                       sec.axis = sec_axis(~ . * stable_mod2,
                                           "Adult wasp abundance")) +
    theme(legend.position = "none")



stab_p2
# stab_p2 + coord_cartesian(xlim = c(0, 100))


stable_mummies2 %>% 
    filter(p_res == "p[res] == 0.5") %>% 
    ggplot(aes(time, N, color = sy)) +
    geom_line()


# ggsave("~/Desktop/stability_plots2.pdf", stab_p, width = 6.5, height = 6.5)


```




### Stability to perturbations

Below shows that the coexistence results can be relatively stable
to various perturbations, especially as adult wasp survival decreases.
We perturbed the susceptible line, resistant line, and adult wasps.
See captions for details.


```{r pert-sims-captions}
fig4_cap <- paste("Aphids and wasps over time, with varying adult",
                  "wasp survival and perturbation types.",
                  "Sub-panel columns separate the adult wasp survival rates used,",
                  "and labels above indicate how they relate",
                  "to the default value ($\\hat{s}_y$).",
                  "Sub-panel rows separate the following perturbation types:",
                  "'none' indicates no perturbation,",
                  "'susceptible' indicates the susceptible line was reduced by 50\\%,",
                  "'resistant' indicates the resistant line was reduced by 50\\%,",
                  "and",
                  "'wasps' indicates the wasps were reduced by 90\\%.",
                  "The black arrows indicate when the perturbations occurred.",
                  "Gray lines are adult wasps, green are resistant aphids,",
                  "and red are susceptible aphids.")
fig5_cap <- paste("Aphids and wasps over time, with varying perturbation types.",
                  "Sub-panels separate the following perturbation types:",
                  "'susceptible' indicates the susceptible line was reduced by 50\\%,",
                  "and",
                  "'resistant' indicates the resistant line was reduced by 50\\%.",
                  "The black arrows indicate when the perturbations occurred.",
                  "Gray lines are adult wasps, green are resistant aphids,",
                  "and red are susceptible aphids.",
                  "All used an adult survival rate of $\\hat{s}_y / 8$.")
```


```{r perturb-sims, fig.height=5, fig.width=6.5, fig.cap=fig4_cap}


# # Each takes ~10 sec
# pert_sims <- list()
# pert_sims[["none"]] <- crossing(.p_res = 0.4,
#                         .s_y = populations$s_y / stable_lvls) %>%
#     pmap(do_stable_sims, max_t = 1000)
# pert_sims[["resistant"]] <- crossing(.p_res = 0.4,
#                        .s_y = populations$s_y / stable_lvls) %>%
#     pmap(do_stable_sims, max_t = 1000,
#          perturb = tibble(when = 500, who = "resistant", how = 0.5))
# pert_sims[["susceptible"]] <- crossing(.p_res = 0.4,
#                        .s_y = populations$s_y / stable_lvls) %>%
#     pmap(do_stable_sims, max_t = 1000,
#          perturb = tibble(when = 500, who = "susceptible", how = 0.5))
# pert_sims[["wasps"]] <- crossing(.p_res = 0.4,
#                        .s_y = populations$s_y / stable_lvls) %>%
#     pmap(do_stable_sims, max_t = 1000,
#          perturb = tibble(when = 500, who = "wasps", how = 0.1))
# saveRDS(pert_sims, "under_constr/pert_sims.rds")

pert_sims <- readRDS("under_constr/pert_sims.rds")


pert_aphids <- map_dfr(names(pert_sims), 
                       function(.z) {
                           map_dfr(pert_sims[[.z]], ~ .x[["aphids"]]) %>%
                               select(-rep, -p_res) %>% 
                               mutate(pert = .z) %>% 
                               stable_fct()
                       })
pert_mummies <- map_dfr(names(pert_sims), 
                        function(.z) {
                            map_dfr(pert_sims[[.z]], ~ .x[["mummies"]]) %>%
                                select(-rep, -p_res) %>% 
                                mutate(pert = .z) %>% 
                                stable_fct()
                        })
pert_wasps <- map_dfr(names(pert_sims), 
                      function(.z) {
                          map_dfr(pert_sims[[.z]], ~ .x[["wasps"]]) %>%
                              select(-rep, -p_res) %>% 
                              mutate(pert = .z) %>% 
                              stable_fct()
                      })


pert_mod <- max(pert_wasps$wasps) / max(pert_aphids$N)

pert_p <- pert_aphids %>%
    ggplot(aes(time, N)) +
    geom_segment(data = pert_wasps %>%
                   filter(pert != "none") %>%
                   distinct(s_y, pert),
               aes(x = 500, xend = 500, 
                   y = max(pert_aphids$N), yend = max(pert_aphids$N) - 2000), 
               arrow = arrow(length = unit(3, "pt"))) +
    geom_line(data = pert_wasps %>%
                  mutate(N = wasps / pert_mod),
              color = "gray70", size = 0.25) +
    geom_line(aes(color = line), size = 0.25) +
    facet_grid(pert ~ s_y, labeller = label_parsed) +
    scale_color_manual(NULL, values = c("chartreuse3", "firebrick")) +
    scale_linetype_manual(values = c(1, 2)) +
    scale_y_continuous("Aphid abundance", #breaks = 0:2 * 4000,
                       sec.axis = sec_axis(~ . * pert_mod,
                                           "Adult wasp abundance")) +
    theme(legend.position = "none")

# pert_p + coord_cartesian(xlim = c(0, 100))

pert_p

```




In the one case where a perturbation caused exclusion of the wasps and 
resistant line, this appears to be a case of timing of the perturbation.
Below, we used only $\hat{s}_y / 5$ for adult wasp survival, then simulated
the perturbations of the susceptible and resistant lines again, but 100 days
later than in the previous figure.
We chose 100 days because that puts the perturbation in the opposite
phase of the fluctuations, when the resistant line is increasing and the
susceptible is decreasing.



```{r perturb-sims2, fig.height=3, fig.width=3, fig.cap=fig5_cap}

.p_time <- 400

# # Takes ~10 sec
# pert_sims2 <- map(c("resistant", "susceptible"),
#                   function(.x) {
#                       do_stable_sims(.p_res = 0.4, .s_y = populations$s_y / 8,
#                                      max_t = 1000,
#                                      perturb = tibble(when = .p_time,
#                                                       who = .x, how = 0.5))
#                   })
# names(pert_sims2) <- c("resistant", "susceptible")
# saveRDS(pert_sims2, "under_constr/pert_sims2.rds")

pert_sims2 <- readRDS("under_constr/pert_sims2.rds")



pert_aphids2 <- map_dfr(names(pert_sims2), 
                       function(.z) {
                           pert_sims2[[.z]][["aphids"]] %>%
                               select(-rep, -p_res, -s_y) %>% 
                               mutate(pert = .z) %>% 
                               stable_fct()
                       })
pert_wasps2 <- map_dfr(names(pert_sims2), 
                      function(.z) {
                          pert_sims2[[.z]][["wasps"]] %>%
                              select(-rep, -p_res, -s_y) %>% 
                              mutate(pert = .z) %>% 
                              stable_fct()
                      })


pert_mod2 <- max(pert_wasps2$wasps) / max(pert_aphids2$N)

pert_p2 <- pert_aphids2 %>% 
    ggplot(aes(time, N)) +
    geom_segment(x = .p_time, xend = .p_time, 
                 y = max(pert_aphids2$N), yend = max(pert_aphids2$N) - 2000,
               arrow = arrow(length = unit(3, "pt"))) +
    geom_line(aes(color = line), size = 0.25) +
    geom_line(data = pert_wasps2 %>%
                  mutate(N = wasps / pert_mod2),
              color = "gray70", size = 0.25) +
    facet_wrap(~ pert, ncol = 1) +
    scale_color_manual(NULL, values = c("chartreuse3", "firebrick")) +
    scale_y_continuous("Aphid abundance", 
                       sec.axis = sec_axis(~ . * pert_mod2,
                                           "Adult wasp abundance")) +
    theme(legend.position = "none")

pert_p2

```



### Stabilizing mechanism

The vector of abundances for the aphid stages at time $t+1$ ($\mathbf{X}(t+1)$) is

$$
\mathbf{X}(t+1) = S(z(t)) \cdot \mathbf{A}(x(t), Y_m(t)) \cdot (\mathbf{L} \, \mathbf{X}(t))
$$

where $S$ gives survival of aphids as it relates to the the total number of aphids $z$,
$\mathbf{A}$ is a vector of the probabilities of the aphid stages surviving parasitism,
$x$ is the total number of unparasitized aphids,
$Y_m$ is the total number of adult wasps,
and
$\mathbf{L}$ is the aphid line's Leslie matrix.



The probability of not being attacked by wasps 
for a given aphid stage ($A_i$) is

$$
A_i = \left( 1 + \frac{a p_i Y_m}{k (h x + 1)} \right)^{-k}
\textrm{.}
$$

Here, $a$ changes the overall attack rate,
$p_i$ is the relative rate on stage $i$ aphids,
$h$ is the handling time, and
$k$ is the aggregation parameter of the negative binomial distribution.


If we focus on one of the sub-panels from the stability simulations and
plot the attack probabilities through time, we can see that it




```{r stable-mechanism, eval=FALSE}

a <- wasp_attack$a
k <- wasp_attack$k
h <- wasp_attack$h
p_i <- wasp_attack$rel_attack[3] / 2
K <- formals(sim_clonewars)$mean_K * 4
A <- function(x, Y_m) (1 + (a * p_i * Y_m) / (k * (h * x + 1)))^(-k)
S <- function(z) 1 / (1 + z / K)

surv_p <- stable_aphids %>%
    filter(s_y == "hat(s)[y] / 6", p_res == "p[res] == 0.4") %>% 
    ggplot(aes(time, N)) +
    geom_hline(yintercept = c(0, max(stable_aphids$N)), color = "gray70") +
    geom_line(data = stable_aphids %>%
                  filter(s_y == "hat(s)[y] / 6", p_res == "p[res] == 0.4") %>% 
                  group_by(s_y, p_res, time) %>%
                  summarize(N = sum(N), .groups = "drop") %>%
                  left_join(stable_wasps, by = c("s_y", "p_res", "time")) %>%
                  mutate(wasp_surv = A(N, wasps)) %>% 
                  mutate(wasp_surv = wasp_surv * max(stable_aphids$N)),
              aes(time, wasp_surv),
              color = "dodgerblue", size = 0.25) +
    geom_line(aes(color = line), size = 0.25) +
    scale_color_manual(NULL, values = c("chartreuse3", "firebrick")) +
    scale_linetype_manual(values = c(1, 2)) +
    scale_y_continuous("Aphid abundance", #breaks = 0:2 * 4000,
                       sec.axis = sec_axis(~ . / max(stable_aphids$N),
                                           "Wasp survival")) +
    theme(legend.position = "none")

# surv_p + coord_cartesian(xlim = c(0, 100))
surv_p


ggsave("~/Desktop/wasp_survival.pdf", surv_p, width = 6, height = 4)


curve(S(x) * A(x, 10), 1, 10e3, ylim = c(0, 0.6), ylab = "survival", xlab = "# aphids")
curve(S(x) * A(x, 100), 1, 10e3, add = TRUE, col = "red")

.n <- 10000
curve((S(.n) * A(.n, 10)) - (S(.n*x) * A(.n*x, 10)), 0, 1, 
      ylim = c(-0.2, 0.2), ylab = "survival", xlab = "proportion resistant")
curve((S(.n) * A(.n, 100)) - (S(.n*x) * A(.n*x, 100)), 0, 1, 
      add = TRUE, col = "red")
abline(h = 0, lty = 2, col = "gray70")





stable_aphids %>%
    pivot_wider(names_from = line, values_from = N) %>% 
    left_join(stable_wasps, by = c("s_y", "p_res", "time")) %>%
    # Benefits conferred to susceptible line by resistant:
    mutate(res_benefits = A(resistant + susceptible, wasps) -
               A(susceptible, wasps)) %>%
    mutate(res_benefits = res_benefits / max(res_benefits) * 
                      max(max(resistant), max(susceptible))) %>% 
    ggplot(aes(time)) +
    geom_line(aes(y = res_benefits), color = "gray70") +
    geom_line(aes(y = resistant), color = "chartreuse3") +
    geom_line(aes(y = susceptible), color = "firebrick") +
    # coord_cartesian(xlim = c(0, 100)) +
    facet_grid(p_res ~ s_y, labeller = label_parsed)


# ppp <- 
stable_aphids %>%
    pivot_wider(names_from = line, values_from = N) %>% 
    left_join(stable_wasps, by = c("s_y", "p_res", "time")) %>%
    filter(wasps > 0) %>% 
    # Benefits conferred to susceptible line by resistant:
    mutate(total = resistant + susceptible,
           res_benefits = A(total, median(wasps)) - A(susceptible, median(wasps)),
           p_res = resistant / total) %>%
    filter(total > 3000) %>%
    ggplot(aes(p_res, res_benefits)) +
    geom_point(shape = 1, alpha = 0.4) +
    ylab("Reduction in wasp mortality\nfor susceptible line") +
    xlab("Proportion of aphids that are resistant") +
    NULL

ppp

# ggsave("~/Desktop/benefits_from_resistant.pdf", ppp, width = 5, height = 3)








# # Playing around with individual combinations:
# z <- do_stable_sims(.p_res = 0, .s_y = populations$s_y / 1000,  max_t = 1000)
# z$mod <- max(z$wasps$wasps) / max(z$aphids$N)
# 
# 
# z$aphids %>%
#     mutate(line = factor(line, levels = c("resistant", "susceptible"))) %>%
#     ggplot(aes(time, N)) +
#     geom_line(data = z$wasps %>%
#                   mutate(N = wasps / z$mod),
#               color = "gray70", size = 0.25) +
#     geom_line(aes(color = line), size = 0.5) +
#     scale_color_manual(NULL, values = c("chartreuse3", "firebrick")) +
#     scale_linetype_manual(values = c(1, 2)) +
#     scale_y_continuous("Aphid abundance", breaks = 0:2 * 4000,
#                        sec.axis = sec_axis(~ . * z$mod,
#                                            "Adult wasp abundance")) +
#     theme(legend.position = "none")


```

